---
title: "Loops & Iteration Mindmaps"
format: html
---

## Lesson 11: across() Function

```{r}
#| echo: false
#| message: false
library(DiagrammeR)

grViz("
digraph Across {
  graph [rankdir = LR]
  
  node [shape = ellipse]
  
  # Central node
  main [label = 'across()', shape = box, style = filled, fillcolor = lightblue]
  
  # All other nodes
  purpose [label = 'Modify multiple\\ncolumns at once']
  by_name [label = '.cols: by name']
  by_select [label = '.cols: select() helpers']
  by_where [label = '.cols: where()']
  by_everything [label = '.cols: everything()']
  in_mutate [label = 'Use in: mutate()']
  in_summarize [label = 'Use in: summarize()']
  in_filter [label = 'Use in: filter()']
  in_arrange [label = 'Use in: arrange()']
  
  # Edges
  main -> purpose
  main -> by_name
  main -> by_select
  main -> by_where
  main -> by_everything
  main -> in_mutate
  main -> in_summarize
  main -> in_filter
  main -> in_arrange
}
")
```

---

## Lesson 12: map() Functions

```{r}
#| echo: false
#| message: false
grViz("
digraph MapFunctions {
  graph [rankdir = LR]
  
  # Global node styles for better rendering
  node [fontname = Helvetica, fontcolor = black]
  
  # Central node - Changed to a lighter color and explicit penwidth
  mapfamily [label = 'map Family', shape = box, style = filled, fillcolor = '#d3f9d8', penwidth = 2]
  
  # All other nodes
  concept [label = 'Apply function\nto each element', shape = ellipse]
  anon [label = 'Anonymous:\nfunction(x)', shape = ellipse]
  map_list [label = 'map(): returns list', shape = ellipse]
  map_df [label = 'map_df(): returns df', shape = ellipse]
  map_chr [label = 'map_chr(): returns char', shape = ellipse]
  map_dbl [label = 'map_dbl(): returns numeric', shape = ellipse]
  walk [label = 'walk(): side effects', shape = ellipse]
  map1 [label = 'map(): 1 input', shape = ellipse]
  map2 [label = 'map2(): 2 inputs', shape = ellipse]
  pmap [label = 'pmap(): multiple inputs', shape = ellipse]
  
  # Edges
  mapfamily -> concept
  mapfamily -> anon
  mapfamily -> map_list
  mapfamily -> map_df
  mapfamily -> map_chr
  mapfamily -> map_dbl
  mapfamily -> walk
  mapfamily -> map1
  mapfamily -> map2
  mapfamily -> pmap
}
")
```


---

## List Operations

```{r}
#| echo: false
#| message: false

grViz("
digraph ListOps {
  graph [rankdir = LR]
  
  node [shape = ellipse]
  
  # Central node
  center [label = 'List Operations', shape = box, style = filled, fillcolor = wheat]
  
  # All other nodes
  double [label = 'x[[i]]: get tibble']
  single [label = 'x[i]: get list']
  pluck [label = 'pluck(): extract nested']
  keep [label = 'keep(): like filter()']
  reduce [label = 'reduce(): like summarize()']
  nest [label = 'nest(): DF with list-col']
  split [label = 'split(): list of DFs']
  
  # Edges
  center -> double
  center -> single
  center -> pluck
  center -> keep
  center -> reduce
  center -> nest
  center -> split
}
")
```

---

## Helper Functions

```{r}
#| echo: false
#| message: false

grViz("
digraph Helpers {
  graph [rankdir = LR]
  
  node [shape = ellipse]
  
  # Central node
  center [label = 'Helper Functions', shape = box, style = filled, fillcolor = lightcoral]
  
  # All other nodes
  set_names [label = 'set_names():\\nname list elements']
  list_files [label = 'list.files():\\nvector of file paths']
  
  # Edges
  center -> set_names
  center -> list_files
}
")
```

## Quick Reference

| Concept | Function | Returns |
|---------|----------|---------|
| **Modify columns** | `across(.cols, .fns)` | Modified data frame |
| **Map basics** | `map(x, f)` | List |
| | `map_df(x, f)` | Data frame |
| | `map_chr(x, f)` | Character vector |
| | `map_dbl(x, f)` | Numeric vector |
| **Side effects** | `walk(x, f)` | (none) |
| **Multiple inputs** | `map2(x, y, f)` | List |
| | `pmap(list, f)` | List |
| **List subsetting** | `x[[i]]` | Element i |
| | `x[i]` | List with element i |
| **List operations** | `keep(x, .p)` | Filtered list |
| | `reduce(x, f)` | Single value |
| **Nesting** | `nest(.data, ...)` | DF with list-column |
| | `split(x, f)` | Named list of DFs |
