---
title: "NYC Airbnb: Demonstrating Key Data Science Concepts"
author: "William Acosta Lora"
date: "2026-03-08"
format: 
  html:
    toc: true
    code-fold: show
    theme: cosmo
execute:
  warning: false
  message: false
---

## Introduction

This project uses NYC Airbnb data to demonstrate four key concepts we have learned these past couple of weeks in class:

1. **Missing Data**: Finding and handling explicit/implicit missing values
2. **Custom Functions**: Writing reusable functions with tidy evaluation
3. **Base R Subsetting**: Using `$`, `[[]]`, and `[]`
4. **Iteration**: For loops and across() for repeated tasks
```{r setup}
library(tidyverse)
library(knitr)

airbnb <- read_csv("data/raw/archive/AB_NYC_2019.csv") |> 
  mutate(last_review = mdy(last_review))
```

---

## Concept 1: Missing Data

### Finding Explicit Missing Values
```{r missing-explicit}
# Count missing values in each column
airbnb |> 
  summarize(
    name = sum(is.na(name)),
    host_name = sum(is.na(host_name)),
    last_review = sum(is.na(last_review)),
    reviews_per_month = sum(is.na(reviews_per_month))
  ) |> 
  kable(caption = "Count of Missing Values")
```

**Observation**: `last_review` and `reviews_per_month` have 10,052 missing values - these are listings that haven't been reviewed yet.

### Handling Missing Values
```{r missing-handle}
# Fill missing reviews_per_month with 0
airbnb_clean <- airbnb |> 
  mutate(reviews_per_month = coalesce(reviews_per_month, 0))

# Verify
tibble(
  before = sum(is.na(airbnb$reviews_per_month)),
  after = sum(is.na(airbnb_clean$reviews_per_month))
) |> 
  kable(caption = "Missing values before and after coalesce()")
```

### Finding Implicit Missing Values
```{r missing-implicit}
# Check if all borough-room_type combinations exist
all_combos <- airbnb_clean |> 
  complete(neighbourhood_group, room_type)

# Find missing combinations
missing_combos <- all_combos |> 
  anti_join(airbnb_clean, by = c("neighbourhood_group", "room_type"))

if (nrow(missing_combos) > 0) {
  missing_combos |> kable()
} else {
  cat(" All borough-room type combinations have listings\n")
}
```

---

## Concept 2: Custom Functions

### Function 1: Calculate Proportion
```{r function-proportion}
# Function to calculate what % of listings meet a condition
calc_proportion <- function(data, condition) {
  data |> 
    summarize(
      n_total = n(),
      n_meeting_condition = sum({{ condition }}, na.rm = TRUE),
      proportion = mean({{ condition }}, na.rm = TRUE)
    )
}

# Example: What % of listings are entire homes?
calc_proportion(airbnb_clean, room_type == "Entire home/apt") |> 
  kable(digits = 3, caption = "Proportion of Entire Homes")

# Example: What % have 100+ reviews?
calc_proportion(airbnb_clean, number_of_reviews >= 100) |> 
  kable(digits = 3, caption = "Proportion with 100+ reviews")
```

### Function 2: Price Summary by Group
```{r function-price-summary}
# Function with embracing and optional grouping
price_summary <- function(data, group_var) {
  data |> 
    group_by({{ group_var }}) |> 
    summarize(
      n = n(),
      min_price = min(price),
      median_price = median(price),
      max_price = max(price),
      .groups = "drop"
    )
}

# By borough
price_summary(airbnb_clean, neighbourhood_group) |> 
  kable(caption = "Price Summary by Borough")

# By room type
price_summary(airbnb_clean, room_type) |> 
  kable(caption = "Price Summary by Room Type")
```

---

## Concept 3: Base R Subsetting

### Using `$`, `[[]]`, and `[]`
```{r base-r-subset}
# Method 1: Extract column with $
brooklyn_prices_dollar <- airbnb_clean$price[airbnb_clean$neighbourhood_group == "Brooklyn"]

# Method 2: Extract column with [[]]
brooklyn_data <- airbnb_clean[airbnb_clean$neighbourhood_group == "Brooklyn", ]
brooklyn_prices_bracket <- brooklyn_data[["price"]]

# Method 3: Tidyverse (for comparison)
brooklyn_prices_tidy <- airbnb_clean |> 
  filter(neighbourhood_group == "Brooklyn") |> 
  pull(price)

# Verify all three methods give same result
tibble(
  method = c("$ operator", "[[ ]] operator", "tidyverse"),
  n_values = c(length(brooklyn_prices_dollar), 
               length(brooklyn_prices_bracket), 
               length(brooklyn_prices_tidy)),
  mean_price = c(mean(brooklyn_prices_dollar), 
                 mean(brooklyn_prices_bracket), 
                 mean(brooklyn_prices_tidy)),
  identical = c(
    identical(brooklyn_prices_dollar, brooklyn_prices_bracket),
    identical(brooklyn_prices_bracket, brooklyn_prices_tidy),
    TRUE
  )
) |> 
  kable(digits = 2, caption = "Comparison of Subsetting Methods")
```

### Using split() and lapply()
```{r base-r-lists}
# Split data by borough into a list
borough_list <- split(airbnb_clean, airbnb_clean$neighbourhood_group)

# Use lapply to calculate mean price for each borough
mean_prices <- lapply(borough_list, function(df) {
  mean(df$price)
})

# Convert to data frame
data.frame(
  borough = names(mean_prices),
  avg_price = unlist(mean_prices)
) |> 
  arrange(desc(avg_price)) |> 
  kable(digits = 2, caption = "Average Price by Borough (using split + lapply)")
```

---

## Concept 4: Iteration

### For Loop with Pre-allocation
```{r iteration-for-loop}
# Goal: Calculate median price for each borough

# Get unique boroughs
boroughs <- unique(airbnb_clean$neighbourhood_group)

# Pre-allocate storage
median_prices <- vector("numeric", length = length(boroughs))
names(median_prices) <- boroughs

# For loop with seq_along
for (i in seq_along(boroughs)) {
  borough_name <- boroughs[i]
  
  # Subset data
  borough_data <- airbnb_clean |> 
    filter(neighbourhood_group == borough_name)
  
  # Calculate and store
  median_prices[i] <- median(borough_data$price)
}

# Display results
data.frame(
  borough = names(median_prices),
  median_price = median_prices
) |> 
  arrange(desc(median_price)) |> 
  kable(caption = "Median Price by Borough (for loop)")
```

### Using across()
```{r iteration-across}
# Calculate mean and median for multiple numeric variables at once
airbnb_clean |> 
  group_by(room_type) |> 
  summarize(
    across(
      .cols = c(price, minimum_nights, availability_365),
      .fns = list(mean = mean, median = median),
      .names = "{.col}_{.fn}"
    ),
    n = n(),
    .groups = "drop"
  ) |> 
  kable(digits = 1, caption = "Summary Statistics by Room Type (using across)")
```

### Using crossing() for Multiple Scenarios

```{r}
# Create multiple analysis scenarios
scenarios <- crossing(
  borough = c("Manhattan", "Brooklyn"),
  expensive = c(TRUE, FALSE)  # expensive = price > $150
)

scenarios |> 
  kable(caption = "Analysis Scenarios")

# Function to analyze one scenario
analyze_scenario <- function(data, borough, expensive) {
  filtered <- data |> 
    filter(neighbourhood_group == borough)
  
  if (expensive) {
    filtered <- filtered |> filter(price > 150)
  }
  
  tibble(
    n_listings = nrow(filtered),
    avg_price = mean(filtered$price)
  )
}

# Run all scenarios with for loop
results <- vector("list", length = nrow(scenarios))

for (i in seq_len(nrow(scenarios))) {
  results[[i]] <- analyze_scenario(
    data = airbnb_clean,
    borough = scenarios$borough[i],
    expensive = scenarios$expensive[i]
  )
}

# Combine results
bind_cols(scenarios, bind_rows(results)) |> 
  kable(digits = 2, caption = "Results for All Scenarios")
```


---

## Summary

This project demonstrated four key concepts:

**Missing Data**
- Found 10,052 explicit missing values in `reviews_per_month`
- Used `coalesce()` to fill with 0
- Used `complete()` and `anti_join()` to find implicit missing values

**Custom Functions**
- Wrote `calc_proportion()` using `{{ }}` for tidy evaluation
- Wrote `price_summary()` with grouped summaries
- Both functions work with different variables

**Base R Subsetting**
- Used `$`, `[[]]`, and `[]` for extraction
- Used `split()` and `lapply()` for list operations
- Verified all methods produce identical results

**Iteration**
- Used for loop with `seq_along()` and pre-allocated storage
- Used `across()` to summarize multiple variables
- Used `crossing()` to create analysis scenarios

**Dataset**: NYC Airbnb Open Data (2019) - 48,895 listings
