---
title: "Brazilian Companies Capital Stock Analysis"
author: "William Acosta"
date: "2026-01-27"
format: 
  html:
    theme: cosmo
    code-fold: true
---

This week's TidyTuesday explores Brazilian companies from the CNPJ (Cadastro Nacional da Pessoa Jurídica) registry. I investigate how capital stock varies across different company characteristics including legal nature, company size, and owner qualifications for the bottom 99% of companies and the top 1%. 

[Click here to jump to Main Findings](#main-findings) with the key visualizations for this TidyTuesday.

# Setup
```{r}
#| message: false
#| warning: false

library(tidyverse)
library(scales)
library(patchwork)

my_colors <- c("blue", "pink", "orange", "green")
```

# Load Data
```{r}
tuesdata <- tidytuesdayR::tt_load(2026, week = 4)

companies <- tuesdata$companies
legal_nature <- tuesdata$legal_nature
qualifications <- tuesdata$qualifications
size <- tuesdata$size
```

# Data Exploration
```{r}
glimpse(companies)
summary(companies$capital_stock)
```

```{r}
# Data Cleaning
companies_filtered <- companies %>% 
  filter(!is.na(company_size))
```

# Distribution Analysis

## Capital Stock Distribution
```{r}
#| fig-width: 12
#| fig-height: 5

(ggplot(data = companies_filtered, aes(x = capital_stock)) +
  geom_histogram(bins = 50, fill = my_colors[1], alpha = 0.8, color = "white") +
  scale_x_log10(labels = label_number(scale = 1e-6, prefix = "R$")) +
  labs(title = "Histogram of Capital Stock",
       x = "Capital Stock (log scale) in Millions",
       y = "Count")) +

(ggplot(data = companies_filtered, aes(x = capital_stock)) +
  geom_density(fill = my_colors[2], alpha = 0.6) +
  scale_x_log10(labels = label_number(scale = 1e-6,  prefix = "R$")) +
  labs(title = "Density Distribution",
       x = "Capital Stock (log scale) in Millions",
       y = "Density")) +

plot_annotation(
  title = "Distribution of Company Capital Stock"
)
```

The distribution is highly right-skewed, with most companies having modest capital stock and a long tail of high-capital firms.

## Capital Stock by Company Size
```{r}
#| fig-width: 12
#| fig-height: 5

(ggplot(data = companies_filtered, aes(x = company_size, y = capital_stock, fill = company_size)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2) +
  scale_y_log10(labels = label_number(scale = 1e-6, prefix = "R$")) +
  scale_fill_manual(values = my_colors) +
  labs(title = "Boxplot by Company Size",
       y = "Capital Stock (log scale) in Millions")) +

(ggplot(data = companies_filtered, aes(x = company_size, y = capital_stock, fill = company_size)) +
  geom_violin(alpha = 0.7) +
  scale_y_log10(labels = label_number(scale = 1e-6, prefix = "R$")) +
  scale_fill_manual(values = my_colors) +
  labs(title = "Violin Plot by Company Size",
       y = "Capital Stock (log scale) in Millions")) +

plot_annotation(
  title = "Capital Stock Distribution by Company Size"
)
```

# Legal Nature and Owner Qualifications
```{r}
# Data Wrangling 
top_legal <- companies_filtered %>%
  filter(!is.na(legal_nature)) %>%
  group_by(legal_nature) %>%
  summarise(
    count = n(),
    total_capital = sum(capital_stock),
    median_capital = median(capital_stock)
  ) %>%
  arrange(desc(total_capital)) %>%
  slice_head(n = 10)

top_qualifications <- companies_filtered %>%
  filter(!is.na(owner_qualification)) %>%
  group_by(owner_qualification) %>%
  summarise(
    count = n(),
    total_capital = sum(capital_stock),
    median_capital = median(capital_stock)
  ) %>%
  filter(count >= 100) %>%
  arrange(desc(median_capital)) %>%
  slice_head(n = 10)
```

```{r}
#| fig-width: 12
#| fig-height: 10

(ggplot(data = top_legal, aes(x = total_capital, y = reorder(legal_nature, total_capital))) +
  geom_col(fill = my_colors[1], alpha = 0.8) +
  scale_x_continuous(labels = label_number(scale = 1e-9, prefix = "R$")) +
  labs(title = "Top 10 Legal Natures by Total Capital",
       x = "Total Capital Stock in Billions",
       y = NULL)) /

(ggplot(data = top_qualifications, aes(x = median_capital, y = reorder(owner_qualification, median_capital))) +
  geom_col(fill = my_colors[3], alpha = 0.8) +
  scale_x_continuous(labels = label_number(scale = 1e-6, prefix = "R$")) +
  labs(title = "Top 10 Owner Qualifications by Median Capital",
       x = "Median Capital Stock in Millions",
       y = NULL))
```

# Top 1% Analysis
```{r}
capital_99th <- quantile(companies_filtered$capital_stock, 0.99)
high_capital <- companies_filtered %>%
  filter(capital_stock >= capital_99th)
```

## Characteristics of Top 1% Companies
```{r}
# Data Wrangling
top1_legal <- high_capital %>%
  count(legal_nature, sort = TRUE) %>%
  slice_head(n = 10)

top1_size <- high_capital %>%
  count(company_size, sort = TRUE)

top1_qual <- high_capital %>%
  count(owner_qualification, sort = TRUE) %>%
  slice_head(n = 8)
```

```{r}
#| fig-width: 12
#| fig-height: 12

(ggplot(data = top1_legal, aes(x = n, y = reorder(legal_nature, n))) +
  geom_col(fill = my_colors[2], alpha = 0.8) +
  labs(title = "Top 1%: Legal Nature",
       x = "Count",
       y = NULL)) /

(ggplot(data = top1_size, aes(x = n, y = reorder(company_size, n))) +
  geom_col(fill = my_colors[1], alpha = 0.8) +
  labs(title = "Top 1%: Company Size",
       x = "Count",
       y = NULL)) /

(ggplot(data = top1_qual, aes(x = n, y = reorder(owner_qualification, n))) +
  geom_col(fill = my_colors[3], alpha = 0.8) +
  labs(title = "Top 1%: Owner Qualification",
       x = "Count",
       y = NULL)) +

plot_annotation(
  title = "Characteristics of Highest-Capital Companies"
)
```
```{r}
# Calculate key statistics for annotation
median_micro <- median(companies_filtered$capital_stock[companies_filtered$company_size == "micro-enterprise"])
median_other <- median(companies_filtered$capital_stock[companies_filtered$company_size == "other"])

pct_micro_below_1M <- mean(companies_filtered$capital_stock[companies_filtered$company_size == "micro-enterprise"] < 1e6) * 100
pct_other_above_10M <- mean(companies_filtered$capital_stock[companies_filtered$company_size == "other"] > 10e6) * 100
```

# Main Findings

## Comparative Analysis: Capital Gap by Company Size
```{r}
#| fig-width: 12
#| fig-height: 6

ggplot(data = companies_filtered, aes(x = capital_stock, fill = company_size)) +
  geom_density(alpha = 0.6) +
  geom_vline(xintercept = capital_99th, linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = capital_99th * 1.5, y = 3.5, 
           label = paste0("99th percentile\n(Top 1% threshold)\nR$", round(capital_99th/1e6, 1), "M"), 
           color = "red", size = 3.5, hjust = 0) +
  scale_x_log10(labels = label_number(scale = 1e-6, prefix = "R$")) +
  scale_fill_viridis_d(
    option = "mako",
    name = "Company Size",
    labels = c("Micro Enterprise", "Other (Large)", "Small Enterprise"),
    begin = 0.2,
    end = 0.8
  ) +
  labs(
    title = "Size Determines Access to Capital: Micro Enterprises Locked Out of Elite Tier",
    subtitle = paste0(round(pct_micro_below_1M, 0), "% of micro enterprises have capital under R$1M, while ", 
                     round(pct_other_above_10M, 0), "% of 'Other' companies exceed R$10M The separation between company size categories reveals \nstructural inequality in Brazilian business. Micro enterprises have the lowest capital levels with no representation above the 99th percentile line. In contrast, \nlarger companies dominate the high-capital space.") ,
    x = "Capital Stock in Millions (log scale)",
    y = "Density",
    caption = "    Companies with capital stock > R$150,000
    Source: CNPJ Open Data (Brazilian Ministry of Finance) | Analysis: William Acosta | TidyTuesday 2026-01-27"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "gray30", margin = margin(b = 10)),
    plot.caption = element_text(hjust = 0, size = 8, color = "gray50", margin = margin(t = 10), lineheight = 1.3),
    legend.position = "bottom"
  )
```

## Extreme Wealth Concentration: Top 1% vs Bottom 99%
```{r}
# Data Wrangling
companies_comparison <- companies_filtered %>%
  mutate(top1_status = if_else(capital_stock >= capital_99th, "Top 1%", "Bottom 99%"))

total_capital <- sum(companies_filtered$capital_stock)
top1_capital <- sum(high_capital$capital_stock)
top1_percentage <- (top1_capital / total_capital) * 100
```

```{r}
#| fig-width: 10
#| fig-height: 6

ggplot(data = companies_comparison, aes(x = capital_stock, fill = top1_status)) +
  geom_density(alpha = 0.6) +
  scale_x_log10(labels = label_number(scale = 1e-6, prefix = "R$")) +
  scale_fill_manual(values = c("Bottom 99%" = "gray60", "Top 1%" = "red")) +
  labs(
    title = "Brazilian Companies: Extreme Capital Concentration in Top 1%",
    subtitle = paste0("The top 1% holds ", round(top1_percentage, 2), "% of total capital, showing minimal overlap with bottom 99%"),
    x = "Capital Stock in Millions (log scale)",
    y = "Density",
    fill = "Company Group",
    caption = "Source: CNPJ Open Data (Brazilian Ministry of Finance) | Analysis: William Acosta | TidyTuesday 2026-01-27"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "gray30", margin = margin(b = 10)),
    plot.caption = element_text(hjust = 0, size = 8, color = "gray50", margin = margin(t = 10), lineheight = 1.3),
    legend.position = "bottom"
  )
```

## Legal Structure and Size: Who Reaches the Top 1%?
```{r}
top_legal_names <- top_legal$legal_nature[1:5]

legal_size_summary <- companies_comparison %>%
  filter(legal_nature %in% top_legal_names) %>%
  mutate(legal_nature = fct_recode(legal_nature,
                                 "LLC" = "Limited Liability Business Company (LLC)",
                                 "PTC" = "Publicly Traded Corporation",
                                 "PHC" = "Privately Held Corporation",
                                 "SIP" = "Simple Innovation Company",
                                 "SP" = "Sole Propriertorship")) %>%
  mutate(company_size = fct_recode(company_size,
                                 "Micro" = "micro-enterprise",
                                 "Small" = "small-enterprise",
                                 "Other" = "other")) %>%
  group_by(legal_nature, company_size) %>%
  summarise(
    median_capital = median(capital_stock),
    pct_top1 = mean(top1_status == "Top 1%") * 100,
    .groups = "drop"
  )
```
```{r}
#| fig-width: 14
#| fig-height: 6

ggplot(data = legal_size_summary, aes(y = median_capital, x = company_size, fill = pct_top1)) +
  geom_col(alpha = 0.9) +
  facet_grid(~ legal_nature) +
  scale_y_log10(labels = label_number(scale = 1e-6, prefix = "R$")) +
  scale_fill_viridis_c(option = "cividis", name = "% in Top 1%", direction = -1) +
  labs(
    title = "Publicly Traded Corporations Dominate the Top 1% of Brazilian Companies",
    subtitle = "Median capital stock by legal structure and company size, colored by top 1% concentration. \nColor shows % of companies in the top 1% by capital. The top 1% holds 99.23% of total capital, concentrated heavily in 'Other'-sized PTCs where over 50%\nof companies are in the elite tier. Micro and small enterprises across all structures remain almost entirely in the bottom 99%, revealing extreme wealth inequality.",
    x = "Company Size",
    y = "Median Capital in Millions (log scale)",
    caption ="    LLC: Limited Liability Company | PHC: Privately Held Corporation | PTC: Publicly Traded Corporation | SIP: Simple Innovation Company | SP: Sole Proprietorship
    Source: CNPJ Open Data (Brazilian Ministry of Finance) | Analysis: William Acosta | TidyTuesday 2026-01-27"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "gray30", margin = margin(b = 10)),
    axis.text.x = element_text(hjust = 1, vjust = 1),
    plot.caption = element_text(hjust = 0, size = 8, color = "gray30", margin = margin(t = 10), lineheight = 1.3),
    legend.position = "right"
  )
```

## Summary

-   The top 1% of companies hold 99.23% of total capital

-   This extreme inequality means the bottom 99% of companies share less than 1% of total capital

-   The "other" (large) size category dominates the highest capital levels

-   Publicly Traded Corporations have the highest concentration in the top 1%, especially in the "Other" size category

-   Owner qualifications did not seem to have too much correlation on company capital.

# References

### 1. Combining Plots with Patchwork

-   Tutorial: https://patchwork.data-imaginist.com/articles/guides/assembly.html
-   Used for: Combining plots with + (side by side) and / (stacked vertically)

### 2. Data Wrangling with dplyr

-  I looked into this book for understanding: all common verbs and count(sort = TRUE) which I did not know prior : https://r4ds.hadley.nz/data-transform
-  Stack Overflow helped me understand reordering factors: https://stackoverflow.com/questions/5208679/order-bars-in-ggplot2-bar-graph
-  Used all these for sorting data, selecting top N rows, and reordering bars with reorder()

### 3. ggplot2 Scales and Aesthetics

-   Scale functions: https://ggplot2.tidyverse.org/reference/scale_continuous.html
-   Manual scales: https://ggplot2.tidyverse.org/reference/scale_manual.html
-   Labels for prefixes since we are using Brazilian currency and logs: https://scales.r-lib.org/reference/label_number.html
-   Stack Overflow helped me look at different ways to use log transformations just for reference but I ended up using the scale_x_log10:      https://stackoverflow.com/questions/14255533/how-to-put-a-ggplot2-plot-on-log-scale.
-  Used for dealing with colors to differentiate plots easier for viewers, dealing with money visualizations since many businesses capitals do not have linear relationships I used a log transformation to better visualize trends. Also used for scale_x_log10(), scale_fill_manual(values = my_colors), custom colors
