---
title: "Brazilian Companies Capital Stock Analysis"
author: "William Acosta"
date: "2026-01-27"
format: 
  html:
    theme: cosmo
    code-fold: true
---

This week's TidyTuesday explores Brazilian companies from the CNPJ (Cadastro Nacional da Pessoa Jurídica) registry.
I investigate how capital stock varies across different company characteristics including legal nature, company size, and owner qualifications.

# Setup
```{r}
#| message: false
#| warning: false

library(tidyverse)
library(scales)
library(patchwork)

my_colors <- c("blue", "pink", "orange", "green")
```

# Load Data
```{r}
tuesdata <- tidytuesdayR::tt_load(2026, week = 4)

companies <- tuesdata$companies
legal_nature <- tuesdata$legal_nature
qualifications <- tuesdata$qualifications
size <- tuesdata$size
```

# Data Exploration
```{r}
glimpse(companies)
summary(companies$capital_stock)
```
```{r}
tibble(
  Metric = c("Number of companies", 
             "Unique legal natures", 
             "Unique owner qualifications", 
             "Size categories"),
  Count = comma(c(nrow(companies),
            n_distinct(companies$legal_nature),
            n_distinct(companies$owner_qualification),
            n_distinct(companies$company_size)))
)
```

# Distribution Analysis

## Capital Stock Distribution
```{r}
#| fig-width: 12
#| fig-height: 5

(ggplot(data = companies, aes(x = capital_stock)) +
  geom_histogram(bins = 50, fill = my_colors[1], alpha = 0.8, color = "white") +
  scale_x_log10(labels = label_number(scale = 1e-6, suffix = "M", prefix = "R$")) +
  labs(title = "Histogram of Capital Stock",
       x = "Capital Stock (log scale)",
       y = "Count")) +

(ggplot(data = companies, aes(x = capital_stock)) +
  geom_density(fill = my_colors[2], alpha = 0.6) +
  scale_x_log10(labels = label_number(scale = 1e-6, suffix = "M", prefix = "R$")) +
  labs(title = "Density Distribution",
       x = "Capital Stock (log scale)",
       y = "Density")) +

plot_annotation(
  title = "Distribution of Company Capital Stock"
)
```

The distribution is highly right-skewed, with most companies having modest capital stock and a long tail of high-capital firms.

## Capital Stock by Company Size
```{r}
#| fig-width: 12
#| fig-height: 5

(ggplot(data = companies %>% filter(!is.na(company_size)), 
        aes(x = company_size, y = capital_stock, fill = company_size)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2) +
  scale_y_log10(labels = label_number(scale = 1e-6, suffix = "M", prefix = "R$")) +
  scale_fill_manual(values = my_colors) +
  labs(title = "Boxplot by Company Size",
       x = NULL,
       y = "Capital Stock (log scale)") +
  theme(legend.position = "none")) +

(ggplot(data = companies %>% filter(!is.na(company_size)), 
        aes(x = company_size, y = capital_stock, fill = company_size)) +
  geom_violin(alpha = 0.7) +
  scale_y_log10(labels = label_number(scale = 1e-6, suffix = "M", prefix = "R$")) +
  scale_fill_manual(values = my_colors) +
  labs(title = "Violin Plot by Company Size",
       x = NULL,
       y = "Capital Stock (log scale)") +
  theme(legend.position = "none")) +

plot_annotation(
  title = "Capital Stock Distribution by Company Size"
)
```
```{r}
companies %>%
  filter(!is.na(company_size)) %>%
  group_by(company_size) %>%
  summarise(
    count = comma(n()),
    median_capital = dollar(median(capital_stock), prefix = "R$"),
    mean_capital = dollar(mean(capital_stock), prefix = "R$"),
    total_capital = dollar(sum(capital_stock), prefix = "R$", scale = 1e-9, suffix = "B")
  )
```

# Legal Nature and Owner Qualifications
```{r}
#| fig-width: 12
#| fig-height: 10

top_legal <- companies %>%
  filter(!is.na(legal_nature)) %>%
  group_by(legal_nature) %>%
  summarise(
    count = n(),
    total_capital = sum(capital_stock),
    median_capital = median(capital_stock)
  ) %>%
  arrange(desc(total_capital)) %>%
  slice_head(n = 10)

top_qualifications <- companies %>%
  filter(!is.na(owner_qualification)) %>%
  group_by(owner_qualification) %>%
  summarise(
    count = n(),
    total_capital = sum(capital_stock),
    median_capital = median(capital_stock)
  ) %>%
  filter(count >= 100) %>%
  arrange(desc(median_capital)) %>%
  slice_head(n = 10)

(ggplot(data = top_legal, aes(x = total_capital, y = reorder(legal_nature, total_capital))) +
  geom_col(fill = my_colors[2], alpha = 0.8) +
  scale_x_continuous(labels = label_number(scale = 1e-9, suffix = "B", prefix = "R$")) +
  labs(title = "Top 10 Legal Natures by Total Capital",
       x = "Total Capital Stock",
       y = NULL)) /

(ggplot(data = top_qualifications, aes(x = median_capital, y = reorder(owner_qualification, median_capital))) +
  geom_col(fill = my_colors[3], alpha = 0.8) +
  scale_x_continuous(labels = label_number(scale = 1e-6, suffix = "M", prefix = "R$")) +
  labs(title = "Top 10 Owner Qualifications by Median Capital",
       x = "Median Capital Stock",
       y = NULL))
```
```{r}
# Tables
top_legal %>%
  mutate(across(c(total_capital, median_capital), ~dollar(.x, prefix = "R$", scale = 1e-9, suffix = "B")),
         count = comma(count))

top_qualifications %>%
  mutate(across(c(total_capital, median_capital), ~dollar(.x, prefix = "R$", scale = 1e-6, suffix = "M")),
         count = comma(count))
```

# Top 1% Analysis
```{r}
capital_99th <- quantile(companies$capital_stock, 0.99)
high_capital <- companies %>%
  filter(capital_stock >= capital_99th)

tibble(
  Metric = c("99th percentile threshold", "Companies in top 1%", "Total capital in top 1%"),
  Value = c(
    dollar(capital_99th, prefix = "R$", scale = 1e-6, suffix = "M"),
    comma(nrow(high_capital)),
    dollar(sum(high_capital$capital_stock), prefix = "R$", scale = 1e-9, suffix = "B")
  )
)
```

## Characteristics of Top 1% Companies
```{r}
#| fig-width: 12
#| fig-height: 12

(high_capital %>%
  count(legal_nature, sort = TRUE) %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = n, y = reorder(legal_nature, n))) +
  geom_col(fill = my_colors[2], alpha = 0.8) +
  labs(title = "Top 1%: Legal Nature",
       x = "Count",
       y = NULL)) /

(high_capital %>%
  count(company_size, sort = TRUE) %>%
  ggplot(aes(x = n, y = reorder(company_size, n))) +
  geom_col(fill = my_colors[1], alpha = 0.8) +
  labs(title = "Top 1%: Company Size",
       x = "Count",
       y = NULL)) /

(high_capital %>%
  count(owner_qualification, sort = TRUE) %>%
  slice_head(n = 8) %>%
  ggplot(aes(x = n, y = reorder(owner_qualification, n))) +
  geom_col(fill = my_colors[3], alpha = 0.8) +
  labs(title = "Top 1%: Owner Qualification",
       x = "Count",
       y = NULL)) +

plot_annotation(
  title = "Characteristics of Highest-Capital Companies"
)
```
