---
title: "Brazilian Companies Capital Stock Analysis"
author: "William Acosta"
date: "2026-01-27"
format: 
  html:
    theme: cosmo
    code-fold: true
---

This week's TidyTuesday explores Brazilian companies from the CNPJ (Cadastro Nacional da Pessoa Jurídica) registry.
I investigate how capital stock varies across different company characteristics including legal nature, company size, and owner qualifications.

# Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(scales)
library(patchwork)

my_colors <- c("blue", "pink", "orange", "green")
```

# Load Data

```{r}
tuesdata <- tidytuesdayR::tt_load(2026, week = 4)

companies <- tuesdata$companies
legal_nature <- tuesdata$legal_nature
qualifications <- tuesdata$qualifications
size <- tuesdata$size
```

# Data Exploration

```{r}
glimpse(companies)
summary(companies$capital_stock)
```

```{r}
# Data Cleaning
companies_filtered <- companies %>% 
  filter(!is.na(company_size))
```

# Distribution Analysis

## Capital Stock Distribution

```{r}
#| fig-width: 12
#| fig-height: 5

(ggplot(data = companies_filtered, aes(x = capital_stock)) +
  geom_histogram(bins = 50, fill = my_colors[1], alpha = 0.8, color = "white") +
  scale_x_log10(labels = label_number(scale = 1e-6, prefix = "R$")) +
  labs(title = "Histogram of Capital Stock",
       x = "Capital Stock (log scale) in Millions",
       y = "Count")) +

(ggplot(data = companies_filtered, aes(x = capital_stock)) +
  geom_density(fill = my_colors[2], alpha = 0.6) +
  scale_x_log10(labels = label_number(scale = 1e-6,  prefix = "R$")) +
  labs(title = "Density Distribution",
       x = "Capital Stock (log scale) in Millions",
       y = "Density")) +

plot_annotation(
  title = "Distribution of Company Capital Stock"
)
```

The distribution is highly right-skewed, with most companies having modest capital stock and a long tail of high-capital firms.

## Capital Stock by Company Size

```{r}
#| fig-width: 12
#| fig-height: 5

(ggplot(data = companies_filtered, aes(x = company_size, y = capital_stock, fill = company_size)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2) +
  scale_y_log10(labels = label_number(scale = 1e-6, prefix = "R$")) +
  scale_fill_manual(values = my_colors) +
  labs(title = "Boxplot by Company Size",
       y = "Capital Stock (log scale) in Millions")) +

(ggplot(data = companies_filtered, aes(x = company_size, y = capital_stock, fill = company_size)) +
  geom_violin(alpha = 0.7) +
  scale_y_log10(labels = label_number(scale = 1e-6, prefix = "R$")) +
  scale_fill_manual(values = my_colors) +
  labs(title = "Violin Plot by Company Size",
       y = "Capital Stock (log scale) in Millions")) +

plot_annotation(
  title = "Capital Stock Distribution by Company Size"
)
```

# Legal Nature and Owner Qualifications

```{r}
# Data Wrangling 
top_legal <- companies_filtered %>%
  filter(!is.na(legal_nature)) %>%
  group_by(legal_nature) %>%
  summarise(
    count = n(),
    total_capital = sum(capital_stock),
    median_capital = median(capital_stock)
  ) %>%
  arrange(desc(total_capital)) %>%
  slice_head(n = 10)

top_qualifications <- companies_filtered %>%
  filter(!is.na(owner_qualification)) %>%
  group_by(owner_qualification) %>%
  summarise(
    count = n(),
    total_capital = sum(capital_stock),
    median_capital = median(capital_stock)
  ) %>%
  filter(count >= 100) %>%
  arrange(desc(median_capital)) %>%
  slice_head(n = 10)

```

```{r}
#| fig-width: 12
#| fig-height: 10
(ggplot(data = top_legal, aes(x = total_capital, y = reorder(legal_nature, total_capital))) +
  geom_col(fill = my_colors[1], alpha = 0.8) +
  scale_x_continuous(labels = label_number(scale = 1e-9, , prefix = "R$")) +
  labs(title = "Top 10 Legal Natures by Total Capital",
       x = "Total Capital Stock in Billions",
       y = NULL)) /

(ggplot(data = top_qualifications, aes(x = median_capital, y = reorder(owner_qualification, median_capital))) +
  geom_col(fill = my_colors[3], alpha = 0.8) +
  scale_x_continuous(labels = label_number(scale = 1e-6, prefix = "R$")) +
  labs(title = "Top 10 Owner Qualifications by Median Capital",
       x = "Median Capital Stock in Millions",
       y = NULL))
```

# Top 1% Analysis

```{r}
capital_99th <- quantile(companies_filtered$capital_stock, 0.99)
high_capital <- companies_filtered %>%
  filter(capital_stock >= capital_99th)
```

## Characteristics of Top 1% Companies

```{r}
# Data Wrangling
top1_legal <- high_capital %>%
  count(legal_nature, sort = TRUE) %>%
  slice_head(n = 10)

top1_size <- high_capital %>%
  count(company_size, sort = TRUE)

top1_qual <- high_capital %>%
  count(owner_qualification, sort = TRUE) %>%
  slice_head(n = 8)
```

```{r}
#| fig-width: 12
#| fig-height: 12
(ggplot(data = top1_legal, aes(x = n, y = reorder(legal_nature, n))) +
  geom_col(fill = my_colors[2], alpha = 0.8) +
  labs(title = "Top 1%: Legal Nature",
       x = "Count",
       y = NULL)) /

(ggplot(data = top1_size, aes(x = n, y = reorder(company_size, n))) +
  geom_col(fill = my_colors[1], alpha = 0.8) +
  labs(title = "Top 1%: Company Size",
       x = "Count",
       y = NULL)) /

(ggplot(data = top1_qual, aes(x = n, y = reorder(owner_qualification, n))) +
  geom_col(fill = my_colors[3], alpha = 0.8) +
  labs(title = "Top 1%: Owner Qualification",
       x = "Count",
       y = NULL)) +

plot_annotation(
  title = "Characteristics of Highest-Capital Companies"
)
```

# Comparative Analysis

```{r}
#| fig-width: 10
#| fig-height: 6
  ggplot(data=companies_filtered, aes(x = capital_stock, fill = company_size)) +
  geom_density(alpha = 0.5) +
  scale_x_log10(labels = label_number(scale = 1e-6, prefix = "R$")) +
  scale_fill_manual(values = my_colors) +
  labs(title = "Capital Stock Density by Company Size",
       x = "Capital Stock in Millions (log scale)",
       y = "Density",
       fill = "Company Size") 
```

# Key Insights

```{r}
total_capital <- sum(companies_filtered$capital_stock)
top1_capital <- sum(high_capital$capital_stock)
top1_percentage <- (top1_capital / total_capital) * 100
top1_percentage

```

# Top 1% VS Median Companies

## Density Comparison

```{r}
# Data Wrangling
companies_comparison <- companies_filtered %>%
  mutate(top1_status = if_else(capital_stock >= capital_99th, "Top 1%", "Bottom 99%"))
```

```{r}
#| fig-width: 10
#| fig-height: 6

# Density comparison
ggplot(data = companies_comparison, aes(x = capital_stock, fill = top1_status)) +
  geom_density(alpha = 0.6) +
  scale_x_log10(labels = label_number(scale = 1e-6, prefix = "R$")) +
  scale_fill_manual(values = c("Bottom 99%" = "gray60", "Top 1%" = "red")) +
  labs(title = "Capital Stock Distribution: Top 1% vs Bottom 99%",
       x = "Capital Stock in Millions (log scale)",
       y = "Density",
       fill = "Group")
```

# Top 1% VS Median Companies' Capital by Legal Nature and Size

```{r}
#| fig-width: 12
#| fig-height: 15
top_legal_names <- top_legal$legal_nature[1:5]

legal_size_summary <- companies_comparison %>%
  filter(legal_nature %in% top_legal_names) %>%
  group_by(legal_nature, company_size) %>%
  summarise(
    median_capital = median(capital_stock),
    pct_top1 = mean(top1_status == "Top 1%") * 100
  )

ggplot(data = legal_size_summary, aes(x = company_size, y = median_capital, fill = pct_top1)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~ legal_nature, ncol = 1) +
  scale_y_log10(labels = label_number(scale = 1e-6, prefix = "R$")) +
  scale_fill_gradient(low = "lightblue", high = "darkred") +
  labs(title = "Median Capital by Legal Nature and Size (colored by % in Top 1%)",
       x = "Company Size",
       y = "Median Capital in Millions (log scale)",
       fill = "% in Top 1%") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

**Summary:**

-   The top 1% of companies hold 99.23% of total capital

-   This extreme inequality means the bottom 99% of companies share less than 1% of total capital

-   The "other" size category have mostly highest capital levels

-   Limited Liability Business Companies dominate both by count and capital

-   Director/Officer qualifications tend to have higher median capital stock

# References

### 1. Combining Plots with Patchwork

-   Tutorial: https://patchwork.data-imaginist.com/articles/guides/assembly.html
-   Used for: Combining plots with + (side by side) and / (stacked vertically)

### 2. Data Wrangling with dplyr

-   I looked into this book for understanding: all common verbs and count(sort = TRUE) which I did not know prior : https://r4ds.hadley.nz/data-transform
-   Stack Overflow helped me understand reordering factors: https://stackoverflow.com/questions/5208679/order-bars-in-ggplot2-bar-graph
-   Used all these for sorting data, selecting top N rows, and reordering bars with reorder()

### 3. ggplot2 Scales and Aesthetics

-   Scale functions: https://ggplot2.tidyverse.org/reference/scale_continuous.html
-   Manual scales: https://ggplot2.tidyverse.org/reference/scale_manual.html
-   Labels for prefixes since we are using Brazilian currency and logs: https://scales.r-lib.org/reference/label_number.html
-   Stack Overflow helped me look at different ways to use log transformations just for reference but I ended up using the scale_x_log10: https://stackoverflow.com/questions/14255533/how-to-put-a-ggplot2-plot-on-log-scale.
-   Used for dealing with colors to differentiate plots easier for viewers, dealing with money visualizations since many businesses capitals do not have linear relationships I used a log transformation to better visualize trends. Also used for scale_x_log10(), scale_fill_manual(values = my_colors), custom colors
